Geometric variational problems

### 2020-11-04 NSF2020/Description/specificaim2_ver2.tex
Rearranging files

\subsection{Specific Aim 2: Efficient high-order numerical algorithm for
large-scale simulation of the HAP model}
\label{subsec:specific_aim_2}
% ----------------------------------------------------------------------
\subsubsection{Overview of the potential theory and boundary integral
equation methods} 
The HAP model requires solving the exterior Dirichlet problem of the
screened Laplace equation and the mobility problem of the Stokes flow at
each time step. If discretizing these equations with stencil-based
numerical methods, the computational domain must be truncated, the
volume must be discretized, and artificial boundary conditions must be
imposed. The artificial boundary conditions introduces an additional
source of error, and discretizing the volume, particularly in three
dimensions, results in an excessively large linear system. Moreover, it
is difficult to obtain high-order discretizations when the boundary
surfaces are irregular. Since we are solving constant-coefficient
elliptic PDEs, we apply boundary integral equation (BIE) that remove the
aforementioned obstacles. BIEs represent the solution as a layer
potential with an unknown density function. Then, the correct density
function is found that matching the layer potential with the prescribed
boundary conditions. By construction, layer potentials satisfy the
underlying PDE and the far-field conditions.

Discretizations of BIEs have several advantages when compared to
stencil-based discretizations. First, the dimension of the problem is
reduced by one since the unknown density function is defined on the
boundary of the geometry. Second, by carefully selecting each layer
potential, the resulting linear system is well-conditioned for any
resolution and iterative methods such as GMRES require a
mesh-independent number of iterations to converge. Third, by using
appropriate quadrature methods, the method converges with spectral or
very high-order accuracy.

Challenges are dense matrices, preconditioners, and near-singular
integration

%The HAP model requires solving the exterior Dirichlet problem of the
%screened Laplace equation and the mobility problem of the Stokes flow
%at each time step. Standard numerical methods such as finite difference
%and finite element methods have to truncate the infinite domain to a
%finite computational domain by imposing certain artificial boundary
%conditions on the truncated domain and need the discretization of the
%whole volume of the truncated domain. The artificial boundary
%conditions are very often inexact, introducing another layer of
%approximation that lowers the accuracy of the solution. The
%discretization of the whole volume is very expensive for three
%dimensional problems, leading to excessively large number of unknowns
%in the linear system. Furthermore, it is rather difficult to obtain
%high-order discretization when the boundary surfaces are irregular.

Potential theory and boundary integral equation (BIE) methods remove the
aforementioned obstacles in a very elegant way. The starting point is
that for constant-coefficient PDEs the fundamental solutions, i.e., the
Green's functions, are readily available. This allows us to represent
the solution via the so-called layer potentials that are the convolution
integrals of a kernel and an unknown density only on the boundary. The
kernel is a linear combination of the Green's function and its
derivatives, which ensures that the representation satifies the
underlying PDE and the condition at infinity automatically. And the
boundary condition at the material interfaces together with the jump
relation of the layer potential leads to a boundary integral equation
for the unknown density.

%The BIE method removes the need of imposing artificial boundary
%conditions for exterior problems and reduces the dimension of the
%problem by one, leading to optimal number of unknowns in the solve
%phase. Very often, one is able to construct the so-called {\it second
%kind integral equation} (SKIE) formulation for these problems, where
%the operator is a sum of the identity operator and a compact operator.
%It is well-known that the condition number of the resulting linear
%system for SKIEs is {\it independent} of the mesh size $h$ or
%equivalently, the number of discretization points. This leads to a
%constant number of iterations when iterative solvers are applied to
%solve the resulting linear system. It further improves the accuracy of
%the solution as the relative error of solution will never increase as
%the number of total discretization points increases. On the other hand,
%the condition number of the linear system from finite difference and
%finite element methods generally increases as a power function of the
%total number of dicretization points, with the exact power depending on
%the underlying spatial dimension and the highest order of the
%differential operator in the PDE. This leads to an increase in the
%number of iterations (and hence the total computational cost) and an
%increase in the relative error of the solution as the number of
%dicretization points increases.

The discretization of the integral operator often leads to a dense
matrix, while finite difference/finite element methods result in sparse
matrices, making matrix-vector product much cheaper to compute.
However,since the invention of the original fast multipole method
(FMM)\cite{fmm5}, there have been many fast
algorithms~\cite{fmm1,fmm2,fmm3,fmm4,fmm6,fmm7,fmm8} that reduce the
cost of matrix-vector product to linear ($O(N)$) or quasilinear
complexity ($O(N \log^kN)$) for dense matrices resulting from the BIE
discretization. Fast direct solvers have also been developed
recently~\cite{fds1,fds2,fds3,fds4,fds5,fds6,fds7,fds8,ho2016cpam2,ho2016cpam1,minden2016,minden2017siammms}.
The development of these fast algorithms have removed a huge hurdle for
the use of boundary integral equations, making the BIE method the method
of choice for constant-coefficient PDEs with linear boundary conditions.
For the exterior Dirichlet problem of the screened Laplace equation, the
SKIE formulation in our previous work~\cite{Fu2018_SIAM} can be readily
extened to the 3D case. For the mobility problem of the Stokes flow in
three dimensions that takes the hydrodynamic effect into account, an
SKIE formulation can be found in~\cite{manasthesis}.

% ----------------------------------------------------------------------
\subsubsection{High-order discretization of surface integrals in three
dimensions}
The practical application of integral equation methods requires the
accurate evaluation of boundary integrals with singular, weakly singular
or nearly singular kernels. There are many numerical quadrature schemes
for dealing with line integrals in two
dimensions~\cite{alpert,kapur,sidi,duffy,bruno1,bruno2,davis_1984,graglia_2008,hackbusch_sauter_1994,
jarvenpaa_2003,khayat_2005,kress_boundary_1991,schwab_1992,
ying_2006,beale1,beale2,goodman_1990, haroldson_1998,
lowengrub_1993,schwab_1992,ggq1,ggq2,ggq3,helsing_2008a,helsing_integral_2009,helsing_tutorial_2012}
and some of them are extremely efficient and can achieve arbitrary high
order.

The high-order quadrature rules for the evaluation of surface integrals
in three dimensions, however, are much less developed than the line
integrals in two dimensions. For example, there are no Gaussian
quadratures for integraing polynomials on a flat triangle, even though
efficient quadratures~\cite{xiao2010cma,vioreanu2014} have been
developed recently for such purpose. For weakly singular or singular
integrals,~\cite{bremer2012jcp,bremer2013jcp} constructed high-order
quadratures for surface integrals on a general triangle,
while~\cite{gimbutas2013sisc} presented a fast algorithm for integrating
$1/r$-type singular integrals for surfaces that are homeomorphic to a
sphere. We plan to study the so-called quadrature by expansion (QBX)
scheme~\cite{klockner2013jcp,qbx2} for the evaluation of both singular
and near-singular surface integrals encountered in the discretization of
BIEs in three dimensions. Conceptually, the idea of the QBX to evaluate
singular, hypersingular and near singular integrals on smooth surfaces
is straightforward. That is, the surface is discretized into smooth
triangles and smooth high-order quadratures are applied to evaluate the
expansion coefficients on all source triangles with the QBX expansion
center placed at a point off the surface. One may then form a suitable
expansion (for example, a Taylor expansion) around that center and
evaluate this expansion back at the target point on the surface (or
close to the surface in the near singular case). Compared to the
competing aforementioned quadrature schemes, the QBX quadrature is
attractive because it offers a clear path for being extended to:
\textbf{(1)} handle any singularity, including hypersingular operators,
\textbf{(2)} be usable with any high-order surface discretization,
\textbf{(3)} generate well-conditioned discrete operators to which
iterative methods such as GMRES~\cite{gmres} can be applied in a
black-box fashion, \textbf{(4)} integrate well with fast algorithms such
as the FMM.

In practice, there are still many issues that need to be resolved. For
example, there are now many variants of QBX including global and local
QBX~\cite{klockner2013jcp,rachh2017jcp}, the target-specific
QBX~\cite{siegel2018jcp}, kernel-independent QBX~\cite{abtin2018bit},
and quadrature by two expansions~\cite{ding2019arxiv}. The coupling of
the QBX and the FMM may also lead to certain instability issues which
may require some changes in the fast multipole
method~\cite{wala2018jcp}. Similar to other quadrature methods, there
have been extensive study on the QBX methods in two dimensions, while
its three dimension
version~\cite{wala2019jcp,af2018sisc,siegel2018jcp,wala2019arxiv} has
not been fully studied and the implementation is even more scarce. We
plan to investigate the accuracy and the convergence order of the
various QBX schemes mentioned above, its coupling with the FMM, parallel
implementation issues for large-scale problems, and the application to
our target problems.

% ----------------------------------------------------------------------
\subsubsection{A hybrid method for close-to-touching Janus particles}
The HAP model for membranes will inevitably lead to the case where some
Janus particles will be very close to each other. Even though QBX can
handle such case accurately, it may lead to excessively large number of
unknowns in order to resolve the unknown densities at those
close-to-touching points. In~\cite{gan2016sisc}, a hybrid scheme is
developed to treat the close-to-touching spheres for the interface
problem of the Laplace equation. The key idea is to combine the analytic
image method and spectrally accurate method of moments, where the
analytic image method will capture the most singular part of the unknown
density at the close-to-touching points. With the aid of the FMM, the
scheme achieves linear complexity with much less number of unknowns on
the material interfaces. We plan to study the hybrid scheme to treat
the close-to-touching case for both the screened Laplace equation and
the mobility problem to further improve the efficiency of the numerical
algorithm for the HAP model.

% ----------------------------------------------------------------------
\subsubsection{Other numerical issues}
Another outstanding numerical issue pertaining to simulating the
self-assembly of amphiphilic particles (such as lipid macromolecles) in
a viscous solvent is the collision between amphiphilic particles. The
hydrophobic attraction potential drives the amphiphilic particles to
move towards each other so to minimize exposure to the solvent. Under
such attraction force the fluid is being squeezed out as particles move
toward each other, and the amphiphilic particles can come into physical
contact in finite time. Such particle collisions in a dense rigid body
suspension is a great challenge and can be a bottleneck in large-scale
simulations. In our previous work a Leonard-Jones potential is used to
keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
steric interaction at short ranges introduces great numerical stiffness
that limit the time step size.

Yan {\it et al.} proposed a collision-resolution algorithm within the
framework of boundary integral methods to resolve the non-smooth
many-body dynamics due to collisions by formulating the particle
collision process as a linear complementarity problem with geometric
`non-overlapping' constraint~\cite{Yan2019}. We propose to treat the
particle collisions based on Yan {\it et al.}'s work after the first two
goals (see above) are achieved. A potential complication in combining
the algorithms for particle collision into our QBX integral formulation
is the computational cost for the QBX to refine meshes on particles in
collision.

Once the collision-resolution algorithm is implemented, we will check
the computing performance first before we embark on incorporating
fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
even at the scales of the amphiphilic particle size in some cases. We
propose to extend the high-order time-integration scheme for a
stochastic differential equation~\cite{fu2015pre} to this system.

>>>>

%\subsection{Specific Aim 2: Functional Materials}
\subsection{Specific Aim 2: Hydrodynamics and ...}
\label{subsec:specific_aim_2}
%The mathematics of this proposal investigates how three-dimensional solids deform under long-range surface tension interaction. The results generalize the physical principles used in the design of functional materials, enabling investigators to simulate sub-micron scale systems that are otherwise impractical to realize experimentally. It supplies new insight into organizational principles in physical and biological sciences. 
%
Over the past decade, there has been an explosion of interest in the fabrication of complex microscopic three-dimensional structures 
\cite{Cho2010}. These fabrication processes utilize the fact that at small scales, capillary forces dominate  van der Waals interactions and thermal noise  \cite{Zhang2017}, enabling the coordinated movement and binding of material subunits. Two prominent fabrication techniques are capillary origami 
\cite{Pandey2011,Leong2007,Reynolds2019} and colloidal self-assembly \cite{Dasgupta2017,Siontorou2017}. Capillary origami uses principles of elastocapilirity wherein elastic solids deform under surface tension \cite{Bico2018,VanHonschoten2010}. In self-assembly of colloids, 
particles exhibit similar thermodynamic behavior of molecular systems, except that the behavior occurs over observable, long time scales \cite{Zhang2017}.

There is inherent intellectual value in fabricating novel three dimensional structures: Living cells construct biological molecules like proteins, membranes and viral capsids using hierarchical assembly \cite{Whitesides2002}. At the same time, elastocapillarity and colloidal self-assembly supply model systems for biological processes, like vesicle adhesion and membrane bound protein interaction \cite{Dasgupta2017}, the hydro-polar model of protein folding \cite{Lau1989,Pandey2011} and viral capsid assembly and \cite{CASPAR1962}. Capillarity can serve as a model for the lock and key mechanism binding between the active site of an enzyme and a substrate molecule \cite{Araujo2018}, or treating viruses like adhered particles 
\cite{Agudo-Canalejo2017,Agudo-Canalejo2015,Deserno2004}, or studying decorated vesicles 
\cite{Auth2009,Weikl2002,Atilgan2007}. Although challenging, mathematically analyzing these model systems gives deep insight into biological structure and function.

%Aside from their intellectual value, high-yield batch production of three-dimensional microstructures has application in functional materials engineering \cite{Syms2003,Leong2010,Crane2013}. In mechanics, for instance, elasticapillarity is a guiding principle lithographic etching of electronics \cite{Bico2018}. Researchers have used self-folding structures in the design of micro containers for encapsulated drug delivery \cite{Fernandes2012} and 3D photovoltaic cells \cite{Guo2009}. 
%In chemical engineering, emulsions stabilize droplets through capillary binding or particulates leading 
%to colloidosomes capturing proteins, vitamins, flavors, living cells \cite{Dinsmore2002}. 
%Microsensors, electronic ink and micromotors \cite{Zhang2017,Yan2016} 
%are potential applications involving Janus-like particles.

\begin{wrapfigure}[14]{r}{3.0in}
\vspace*{+15pt}
\centerline{\includegraphics[width=3.0in]{figures/SA3_fig2.pdf}}
\vspace*{5pt}
\caption{{\footnotesize Illustration of self-assembly of hydrophobically labeled faces.
The thick triangles have a rim-hydrophobic label on one side, and are hydrophilic on the other side.
These shapes would potentially aggregate into a tetrahedral container. }}
\label{fig:illustration}
\end{wrapfigure}
The HAP theory is ideally suited to the problem of creating microstructures through capillary action. Specifically, the modeling application considers the assembly of polyhedra from smaller subunits. Our developments, which include viscous interactions, thermal fluctuation, and a long range description of capillarity, will pave the way for dynamical understanding to organizational principles in viral capsid formation, for instance \cite{CASPAR1962,Prasad2012}.

In order to illustrate the novelty of our approach, it is helpful to first describe how current experimental and mathematical techniques deal with polyhedral assembly. Capillary origami uses lithography to etch faces out of a thin, elastic sheet \cite{Pandey2011,Reynolds2019}. The collection of faces form a so-called net, and this net self-folds into a three-dimensional polyhedron (Figure~\ref{fig:illustration}). The motivation is to create micro containers to mimc protein folding \cite{Reynolds2019}. Prediction of geometric structure from its subunits is non-trivial. That is because the final morphology depends on the net structure, physical interactions, and the surrounding media, and researchers have formulated optimality criteria for these nets \cite{Araujo2018,Pandey2011}.

The experimental shapes are hundreds of microns in diameter. Thus, as a key distinction with HAP approach, the driving force for folding in capillary origami and related processes comes from surface energy of the hinges (e.g. soldered hinges that supply a surface tension once the system is heated). Mathematicians have numerically and analytically described the behavior of these surface tension driven hinges \cite{Bico2018,Peraud2014,Brubaker2016}, making it possible to simulate the mechanical problem, in principle, using slender body theory for plates.

To study this problem using HAP theory, we form a collection of thin, prism-like particles in Figure \ref{fig:illustration}. 
Each of these particles serves as a face in the final morphology. The edges of the prisms are labeled hydrophobic according to the boundary condition in the screened Laplace equation boundary value problem. This way, the edges coalesce, forming the desired structures. Furthermore, 
by manipulating the boundary condition in HAP, it is possible to have faces attach and detach with precision, and form a net or other structures if so desired. Our HAP theory therefore generalizes submillimeter theoretical and experimental work to the micron and nanometer sizes.

%There is also the possibility to explore new classes of mathematical materials, like wrinkling in polystyrene sheets \cite{Paulsen2016}, that include the interaction between an elastic material, the viscous solvent, and long range hydrophobic (capillary) interactions. In the case of rigid particles and subunits, the rigid body motions come from the so-called mobility boundary value problem for the Stokes equation. Here, the hydrophobic stress balances with the viscous stress coming from the rigid body motion of the particles in the surrounding solvent. If the particles are deformable, like a protein \cite{Elms2012} or as in the capillary induced bending of rods \cite{Schulman2017}, then we can formulate an elastic energy for deformation. These elastic energies account for surface bending or even permanent charge. To deploy this approach, it is a matter of solving the bodies' momentum balance equations coupled to the fluid through the viscous, elastic, and hydrophobic stress balance.

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/SA3_fig1.pdf}
\end{center}
\caption{(A)--(D) shows the spontaneous insertion of two hydrophobic particles into a vesicle bilayer. 
The driving force is long-range interaction between the hydrophobic, magenta particles and
the bilayer core.  The particle entrainment cause the vesicle to break apart, analogously to surfactant
induced lysis.}
\end{figure}
Finally, further applications of our modeling approach include colloidal interaction on interfaces. Here, the experimental setups use Janus particles to model, for example, membrane bound proteins \cite{Kumar2013,Lehle2008,Zhang2017}. The idea of an amphiphilic Janus particle was introduced by P. G. de Gennes in 1991 \cite{Zhang2017}. The Investigators have significant experience as Janus particles figure heavily into our particle based description of bilayers, and can adjust our modeling approach to account for proteins. This is done by consider mixtures, consisting of many (hundreds to thousands) long, elongated amphiphilic particles for the lipid phase. Larger particles with hydrophobic patches or stripes model the proteins \cite{Jurasek2017}.

The distinction between macroscopic capillarity and microscopic hydrophobic force are relevant here as well.  Prior models use surface energies to account for interaction between particles in an interface \cite{Dasgupta2017}. 
However, this may be missing important physics, and here our modeling approach has the advantage in that protein insertion, for instance, is a self-consistent consequence of the formulation, versus an assumption imposed by boundary conditions.  
The simulations will enable us address important biological questions on how a collection of 
interface bound colloids interact \cite{Yao2013,Salib2013}, 
and how configurations dictate 
macroscopic surface properties and sense background curvature \cite{Cavallaro2011,Liu2018,Sharifi-Mood2015}.
